---
title: "Bayesian Sample Size Estimation for Multilevel Trials"
author: "Ulrich Lösener"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

# Simulation study for the project: Bayesian Sample Size Determination for Multilevel Trials

In this simulation study, I generate multilevel data for N individuals at n timepoints. The within variable is "time" and the between variable "treatment".

First, I call the necessary libraries.

```{r Packages, message=FALSE}
library(tidyr)       # pipes
library(dplyr)       # pipes
library(ggplot2)     # plots
library(nlme)
library(lme4)        # fit multilevel model
library(lmerTest)    # get p-values of multilevel model
library(conflicted)  # conflicting commands from different packages
library(merDeriv)    # extracting vcov matrices (vcov.lmerMod)
library(mgcv)        # extracting vcov matrices
library(bain)        # Bayesian estimation
library(MASS)        # multinorm - already included in lme4?
library(MuMIn)       # for R squared with r.squaredGLMM(inter)

conflicts_prefer(lme4::lmer) # tell R to prefer lme4 over lmer
conflicts_prefer(dplyr::select)
```

```{r Data Generation}
start <- Sys.time()



true.hyp <- 1 # which hypothesis is true? H1: b>0; H2: b=0
n.steps <- 20 # number of different Ns to evaluate 


mbf12 <- numeric(n.steps)
mbf21 <- numeric(n.steps)
mbf.u <- numeric(n.steps)
mbf.c <- numeric(n.steps)
mpmp.a <- numeric(n.steps)
mpmp.b <- numeric(n.steps)
mpmp.c <- numeric(n.steps)

medbf12 <- numeric(n.steps)
quantbf12 <- numeric(n.steps)
prop.bf12 <- numeric(n.steps)
medbf21 <- numeric(n.steps)
medbf.u <- numeric(n.steps)
medbf.c <- numeric(n.steps)
medpmp.a <- numeric(n.steps)
medpmp.b <- numeric(n.steps)
medpmp.c <- numeric(n.steps)

#set.seed(1234)      # for reproducibility
seeds <- vector("list", n.steps) 

#while (quantbf12<.8) {
  
  
  for(j in 1:n.steps){
    
    N <- j*5        # number of subjects
    d <- c(0,1,2,3,4) # time of measurements
    n <- length(d)    # number of measurements per subject
    
    sigmasq.u0 <- 1   # variance of individual deviation from treatment intercept 
    sigmasq.u1 <- 1   # variance of individual deviation from treatment slope
    sigma.u0.u1 <- 0  # covariance between sigmasq.u0 and sigmasq.u1. If positive, then individuals with higher(lower) initial values tend to have a higher (lower) rate of change over time.
    sigmasq.e <- 1    # error variance
    
    # create data vectors
    y <- rep(NA, N)   # data storage
    t <- rep(d, N)
    id <- rep(seq_len(N), each=n)
    treat <- rep(rbinom(N,1,.5), each=n)
    dat0 <- data.frame(id, treat, t)
    
    beta0 <- rep(0, N*n) # average y at t0 for x=0
    beta1 <- rep(0, N*n) # average increase for x=0
    beta2 <- ifelse(true.hyp==1, rep(1, N*n), rep(0, N*n)) # average difference in slopes between conditions
    
    eff.size <- beta2/sqrt(sigmasq.u1)
    
    m <- 1000 # number of datasets
    sets <- vector("list", m) # storage for datasets
    est <- matrix(c(seq(1:m)), nrow = m, ncol = 2) # storage for fixed estimates
    colnames(est) <- c("dataset", "t:treat")
    
    bf12 <- numeric(m) # storage for BFs
    bf21 <- numeric(m)
    bf.u <- numeric(m) # storage for BFs
    bf.c <- numeric(m) # storage for BFs
    pmp.a <- numeric(m) # storage for BFs
    pmp.b <- numeric(m) # storage for BFs
    pmp.c <- numeric(m) # storage for BFs
  
    for (i in 1:m) {
      seeds[[i]] <- .Random.seed
      multinorm <- mvrnorm(n=N, mu=c(0,0), matrix(c(sigmasq.u0, sigma.u0.u1, sigma.u0.u1, sigmasq.u1), nrow=2, ncol=2)) # draw individual deviation from treatment intercept and slope from a multivariate normal distribution with mean 0.
      u0 <- rep(multinorm[,1], each=n)
      u1 <- rep(multinorm[,2], each=n)
      e <- rnorm(N*n, 0, sqrt(sigmasq.e))
      
      y <- beta0 + u0 + beta1*t + beta2*treat*t + u1*t + e
      dat <- data.frame(dat0, y)
      sets[[i]] <- dat
      
      inter <- lme(y ~ t + t:treat, random =~ t - 1 | id, data = dat)
      #, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
      est[i,2] <- inter$coefficients$fixed[3]
      sig <- list(as.matrix(inter$varFix[3,3]))
      
      result <- bain(est[i,2], hypotheses <- "t:treat>0;t:treat=0", n=N, Sigma = sig, group_parameters = 1, joint_parameters = 0)
      
      bf12[i] <- result$BFmatrix[1,2]
      bf21[i] <- result$BFmatrix[2,1]
      bf.u[i] <- result$fit$BF.u[1]
      bf.c[i] <- result$fit$BF.c[1]
      pmp.a[i] <- result$fit$PMPa[1]
      pmp.b[i] <- result$fit$PMPb[1]
      pmp.c[i] <- result$fit$PMPc[1]
    }
     
    mbf12[j] <- mean(bf12)
    mbf21[j] <- mean(bf21)
    mbf.u[j] <- mean(bf.u)
    mbf.c[j] <- mean(bf.c)
    mpmp.a[j] <- mean(pmp.a)
    mpmp.b[j] <- mean(pmp.b)
    mpmp.c[j] <- mean(pmp.c)
    
    medbf12[j] <- median(bf12)
    quantbf12[j] <- quantile(bf12, probs=0.2)
    prop.bf12[j] <- length(bf12[bf12>3])/m
    medbf21[j] <- median(bf21)
    medbf.u[j] <- median(bf.u)
    medbf.c[j] <- median(bf.c)
    medpmp.a[j] <- median(pmp.a)
    medpmp.b[j] <- median(pmp.b)
    medpmp.c[j] <- median(pmp.c)
  }

#}
  
print(Sys.time() - start) # 2 mins for N=10-100 and m=1000
```



```{r}
# results in data frame


plot(quantbf12, type="l")
simres <- data.frame(samplesize,mbf12,mbf21,mbf.u,mbf.c,mpmp.a,mpmp.b,mpmp.c)
ss.seq <- seq(5, N, 5)
beta.seq <- seq(.1, 1, .1)

#ggplot(data=simres) + geom_line(mapping=aes(x=samplesize, y=mbf12, col="blue")) + geom_line(mapping=aes(x=samplesize, y=mbf21, col="red"))  

par(mfrow=c(2,2))
plot(x=ss.seq, y=mbf12, type="l", xlab="N", ylab="mean BF12")
plot(x=ss.seq, y=mbf21, type="l", xlab="N", ylab="mean BF21")
plot(x=ss.seq, y=mbf.u, type="l", xlab="N", ylab="mean BF unconstr.")
plot(x=ss.seq, y=mbf.c, type="l", xlab="N", ylab="mean BF complement")

par(mfrow=c(2,2))
plot(x=ss.seq, y=medbf12, type="l", xlab="N", ylab="median BF12")
plot(x=ss.seq, y=medbf21, type="l", xlab="N", ylab="median BF21")
plot(x=ss.seq, y=medbf.u, type="l", xlab="N", ylab="median BF unconstr.")
plot(x=ss.seq, y=medbf.c, type="l", xlab="N", ylab="median BF complement")

par(mfrow=c(2,2))
plot(x=ss.seq, y=mpmp.a, type="l", xlab="N", ylab="mean PMPa")
plot(x=ss.seq, y=mpmp.b, type="l", xlab="N", ylab="mean PMPb")
plot(x=ss.seq, y=mpmp.c, type="l", xlab="N", ylab="mean PMPc")
plot(x=ss.seq, y=prop.bf12, type="l", xlab="N", ylab="proportion of BFs larger 3")
abline(h=.8, col="red")

```



```{r PDFs}
pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/mBF_N_bet=1.pdf",   
    width = 6, # The width of the plot in inches
    height = 6) # The height of the plot in inches

par(mfrow=c(2,2))
plot(x=seq(10,100,10), y=mbf12, type="l", xlab="N", ylab="mean BF12")
plot(x=seq(10,100,10), y=mbf21, type="l", xlab="N", ylab="mean BF21")
plot(x=seq(10,100,10), y=mbf.u, type="l", xlab="N", ylab="mean BF unconstr.")
plot(x=seq(10,100,10), y=mbf.c, type="l", xlab="N", ylab="mean BF complement")

dev.off()


pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/medBF_N.pdf",   
    width = 6, # The width of the plot in inches
    height = 6) # The height of the plot in inches

par(mfrow=c(2,2))
plot(x=ss.seq, y=medbf12, type="l", xlab="N", ylab="median BF12")
plot(x=ss.seq, y=medbf21, type="l", xlab="N", ylab="median BF21")
plot(x=ss.seq, y=medbf.u, type="l", xlab="N", ylab="median BF unconstr.")
plot(x=ss.seq, y=medbf.c, type="l", xlab="N", ylab="median BF complement")

dev.off()


pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/PMPs_N_bet=1.pdf",   
    width = 6, # The width of the plot in inches
    height = 6) # The height of the plot in inches

par(mfrow=c(2,2))
plot(x=seq(10,100,10), y=mpmp.a, type="l", xlab="N", ylab="mean PMPa")
plot(x=seq(10,100,10), y=mpmp.b, type="l", xlab="N", ylab="mean PMPb")
plot(x=seq(10,100,10), y=mpmp.c, type="l", xlab="N", ylab="mean PMPc")
plot(x=ss.seq, y=prop.bf12, type="l", xlab="N", ylab="Proportion of BFs larger than 3")
abline(h=.8, col="red")

dev.off()


pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/Prop.BFs3_N_H0.pdf",   
    width = 6, # The width of the plot in inches
    height = 6) # The height of the plot in inches

plot(x=ss.seq, y=prop.bf12, type="l", xlab="N", ylab="proportion of BFs larger 3")
abline(h=.8, col="red")

dev.off()

```


The resulting Bayes Factors tell us that $H_1$ is the hypothesis most likely to be true out of the set of considered hypotheses. We therefore conclude that in the treatment group ($X_1$), the symptom level increases more over time compared to the control group ($X_0$).

## Descriptives/plots

```{r Descriptives, echo=F}
# plot individual trajectories
ggplot(data = dat, aes(x = t, y = y, group = id, color = as.factor(treat))) + geom_line()

boxplot(formula = y ~ treat, data = dat) # in a boxplot
```

As can be seen from the figure, individuals differ in their intercept and slope. Subjects in the treatment group ($X_1$) seem to have systematically higher slopes as compared to those in the control group ($X_0$), suggesting an interaction effect between time and treatment. Also, the variability of y scores seems to increase over time, rendering the assumption of compound symmetry untenable. 


```{r Plots, echo=F}
# one line for everyone
ggplot(data = dat, aes(x = t, y = y), color = treat) + 
  geom_jitter(aes(color = treat), width = .1, height = 0) + 
  geom_smooth(method = "lm", formula = 'y ~ x')
# base
p <- ggplot(data = dat, aes(x = t, y = y, group = id, color = treat)) + geom_jitter(width = .1, height = 0)
# different lines for treat
p  + geom_smooth(group = 0, method = "lm", data = subset(dat, dat$treat == 0), formula = 'y ~ x') + geom_smooth(group = 0, method = "lm", data = subset(dat, dat$treat == 1), formula = 'y ~ x')

```

## Intercept only model

```{r Intercept only}
int.only <- lmer(y ~ 1 + (1 | id), data = dat)
summary(int.only) 
logLik(int.only)           
performance::icc(int.only) 
p + geom_smooth(method = "lm", formula = y ~ 1, se = F)
```

--------------------------------------------------------------------------------

## Appendix: In-between models

### Level 1 predictor: time

```{r Lvl 1}
lvl1 <- lmer(y ~ time + (1 | id), data = dat)
summary(lvl1)
logLik(lvl1)           
anova(int.only, lvl1)  
performance::icc(lvl1) 
```


### Level 2 predictor: treat

```{r Lvl 2}
lvl2 <- lmer(y ~ time + treat + (1 | id), data = dat)
summary(lvl2)
logLik(lvl2)           # -1063.76 (df=5) - better fit compared to lvl1 predictor only
anova(lvl1, lvl2)      # likelihood ratio test significant
performance::icc(lvl2) # 0.044
vcov.lmerMod(lvl2)
```

### Random slope for time

```{r Random slope}
rand.slop <- lmer(y ~ time + treat + (1 + time | id), data = dat)
summary(rand.slop)
logLik(rand.slop) 
anova(lvl2, rand.slop)
performance::icc(rand.slop)
```
