---
title: "Bayesian Sample Size Estimation for Multilevel Trials"
author: "Ulrich Lösener"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

# Simulation study for the project: Bayesian Sample Size Determination for Multilevel Trials

In this simulation study, I generate multilevel data for N individuals at n timepoints. The within variable is "time" and the between variable "treatment".

First, I call the necessary libraries.

```{r Packages, message=FALSE}
library(tidyr)       # pipes
library(dplyr)       # pipes
library(ggplot2)     # plots
library(nlme)
library(lme4)        # fit multilevel model
library(lmerTest)    # get p-values of multilevel model
library(conflicted)  # conflicting commands from different packages
library(merDeriv)    # extracting vcov matrices (vcov.lmerMod)
library(mgcv)        # extracting vcov matrices
library(bain)        # Bayesian estimation
library(MASS)        # multinorm - already included in lme4?
library(MuMIn)       # for R squared with r.squaredGLMM(inter)

conflicts_prefer(lme4::lmer) # tell R to prefer lme4 over lmer
conflicts_prefer(dplyr::select)
```

```{r Data Generation}
start <- Sys.time()

mbf12 <- numeric(10)
mbf.u <- numeric(10)
mbf.c <- numeric(10)
mpmp.a <- numeric(10)

for(j in 1:10){
N <- j*10          # number of subjects
d <- c(0,1,2,3,4) # time of measurements
n <- length(d)    # number of measurements per subject

sigmasq.u0 <- 1  # variance of individual deviation from treatment intercept 
sigmasq.u1 <- 1  # variance of individual deviation from treatment slope
sigma.u0.u1 <- 0 # covariance between sigmasq.u0 and sigmasq.u1. If positive, then individuals with higher(lower) initial values tend to have a higher (lower) rate of change over time.
sigmasq.e <- 1  # error variance

# create data vectors
y <- rep(NA, N)  # data storage
t <- rep(d, N)
id <- rep(seq_len(N), each=n)
treat <- rep(c(rep(0, n), rep(1, n)), N*n)
dat0 <- data.frame(id, treat, t)

beta0 <- rep(0, N*n) # average y at t0 for x=0
beta1 <- rep(1, N*n) # average increase for x=0
beta2 <- rep(.2, N*n) # average difference in slopes between conditions

#set.seed(123)      # for reproducibility

m <- 100          # number of datasets
sets <- vector("list", m)     # storage for datasets
est <- matrix(c(seq(1:m)), nrow = m, ncol = 2) # storage for fixed estimates
colnames(est) <- c("dataset", "t:treat")
bf12 <- numeric(m) # storage for BFs
bf.u <- numeric(m) # storage for BFs
bf.c <- numeric(m) # storage for BFs
pmp.a <- numeric(m) # storage for BFs

for (i in 1:m) {
  
  multinorm <- mvrnorm(n=N, mu=c(0,0), matrix(c(sigmasq.u0, sigma.u0.u1, sigma.u0.u1, sigmasq.u1), nrow=2, ncol=2)) # draw individual deviation from treatment intercept and slope from a multivariate normal distribution with mean 0.
  u0 <- rep(multinorm[,1], each=n)
  u1 <- rep(multinorm[,2], each=n)
  e <- rnorm(N*n, 0, sqrt(sigmasq.e))
  
  y <- beta0 + u0 + beta1*t + beta2*treat*t + u1*t + e
  dat <- data.frame(dat0, y)
  sets[[i]] <- dat
  
  inter <- lme(y ~ t + t:treat, random =~ t - 1 | id, data = dat)
  #, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
  est[i,2] <- inter$coefficients$fixed[3]
  sig <- list(as.matrix(inter$varFix[3,3]))
  
  result <- bain(est[i,2], hypotheses <- "t:treat>0;t:treat<0", n=N, Sigma = sig, group_parameters = 1)
  
  bf12[i] <- result$BFmatrix[1,2]
  bf.u[i] <- result$fit$BF.u[1]
  bf.c[i] <- result$fit$BF.c[1]
  pmp.a[i] <- result$fit$PMPa[1]

}

mbf12[j] <- mean(bf12)
mbf.u[j] <- mean(bf.u)
mbf.c[j] <- mean(bf.c)
mpmp.a[j] <- mean(pmp.a)

}
print( Sys.time() - start ) # 1.5 minutes for N=100, n=5, and m=10


pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/mBF12_N.pdf",   
    width = 4, # The width of the plot in inches
    height = 4) # The height of the plot in inches

plot(x=seq(10,100,10), y=mbf12, type="l", xlab="N")

dev.off()


pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/mBFu_N.pdf",   
    width = 4, # The width of the plot in inches
    height = 4) # The height of the plot in inches

plot(x=seq(10,100,10), y=mbf.u, type="l", xlab="N")

dev.off()


plot(x=seq(10,100,10), y=mpmp.a, type="l", xlab="N")


# par(mfrow=c(2,2))
# plot(bf.u, type="l")
# #plot(sort(bf.c), type="l")
# plot(bf12, type="l")
# 
# hist(bf.u, breaks=10)
# #hist(bf.c)
# hist(bf12, breaks=10)
# 
# #hist(pmp.b)
# quantile(bf12, .20)
# 
# # check the bias in estimates of beta 2
# mean(est[,2] - beta2) 



```


The resulting Bayes Factors tell us that $H_1$ is the hypothesis most likely to be true out of the set of considered hypotheses. We therefore conclude that in the treatment group ($X_1$), the symptom level increases more over time compared to the control group ($X_0$).

## Descriptives/plots

```{r Descriptives, echo=F}
# plot individual trajectories
ggplot(data = dat, aes(x = t, y = y, group = id, color = as.factor(treat))) + geom_line()

boxplot(formula = y ~ treat, data = dat) # in a boxplot
```

As can be seen from the figure, individuals differ in their intercept and slope. Subjects in the treatment group ($X_1$) seem to have systematically higher slopes as compared to those in the control group ($X_0$), suggesting an interaction effect between time and treatment. Also, the variability of y scores seems to increase over time, rendering the assumption of compound symmetry untenable. 


```{r Plots, echo=F}
# one line for everyone
ggplot(data = dat, aes(x = t, y = y), color = treat) + 
  geom_jitter(aes(color = treat), width = .1, height = 0) + 
  geom_smooth(method = "lm", formula = 'y ~ x')
# base
p <- ggplot(data = dat, aes(x = t, y = y, group = id, color = treat)) + geom_jitter(width = .1, height = 0)
# different lines for treat
p  + geom_smooth(group = 0, method = "lm", data = subset(dat, dat$treat == 0), formula = 'y ~ x') + geom_smooth(group = 0, method = "lm", data = subset(dat, dat$treat == 1), formula = 'y ~ x')

```

## Intercept only model

```{r Intercept only}
int.only <- lmer(y ~ 1 + (1 | id), data = dat)
summary(int.only) 
logLik(int.only)           
performance::icc(int.only) 
p + geom_smooth(method = "lm", formula = y ~ 1, se = F)
```

--------------------------------------------------------------------------------

## Appendix: In-between models

### Level 1 predictor: time

```{r Lvl 1}
lvl1 <- lmer(y ~ time + (1 | id), data = dat)
summary(lvl1)
logLik(lvl1)           
anova(int.only, lvl1)  
performance::icc(lvl1) 
```


### Level 2 predictor: treat

```{r Lvl 2}
lvl2 <- lmer(y ~ time + treat + (1 | id), data = dat)
summary(lvl2)
logLik(lvl2)           # -1063.76 (df=5) - better fit compared to lvl1 predictor only
anova(lvl1, lvl2)      # likelihood ratio test significant
performance::icc(lvl2) # 0.044
vcov.lmerMod(lvl2)
```

### Random slope for time

```{r Random slope}
rand.slop <- lmer(y ~ time + treat + (1 + time | id), data = dat)
summary(rand.slop)
logLik(rand.slop) 
anova(lvl2, rand.slop)
performance::icc(rand.slop)
```
