---
title: "Bayesian Sample Size Estimation for Multilevel Trials"
author: "Ulrich Lösener"
date: "13-2-2023"
output: html_document
---

# Simulation study for the project: Bayesian Sample Size Determination for Multilevel Trials

In this simulation study, I generate multilevel data for N individuals at n timepoints. The within variable is "time" and the between variable "gender". To make this data appropriate for multilevel models, I introduce some autocorrelation within subjects, meaning that responses from the same subject are more similar than responses from different subjects. 

First, we call the nessecary libraries.

```{r Packages, message=FALSE}
library(tidyr)       # pipes
library(dplyr)       # pipes
library(ggplot2)     # plots
library(lme4)        # fit multilevel model
library(lmerTest)    # get p-values of multilevel model
library(conflicted)  # conflicting commands from different packages
library(faux)        # simulate multilevel data
library(merDeriv)    # extracting vcov matrices (vcov.lmerMod)
library(longitudinal)# for unequal spacing
library(mgcv)        # extracting vcov matrices
library(bain)        # Bayesian estimation

conflict_prefer("lmer", "lmerTest") # so that lmer commands are preferred over lmerTest commands
```

We set the hyperparameters N (number of subjects) and n (number of timepoints)

```{r Set hyperparameters}
N <- 100  # number of subjects
n <- 10   # number of timepoints

# correlation matrix between tpoints with decreasing r for more distant tpoints
cmat <- abs(row(diag(n)) - col(diag(n)))
x <- 0:n
sequ.corr <- .6 * exp(-.8*x)
#plot(x, sequ.corr, type = "l", xlim = c(0,5), ylim = c(0,1))
for (i in 1:n) {
  cmat[cmat==i] <- sequ.corr[i]
}
diag(cmat) <- 1
```

## Data generation

### Version 1

First version of data generation: ICC not specified -- only the timepoint vectors are correlated, no specific intrasubject correlation
```{r Generate data V1}
set.seed(123)

dat0 <- rnorm_multi(n=N/2, vars=n, mu=seq(1, (n), length=n), sd=2, r=cmat)
dat0$gender <- 0
dat1 <- rnorm_multi(n=N/2, vars=n, mu=seq(1, (n/4), length=n), sd=2, r=cmat)
dat1$gender <- 1
dat.wide <- rbind(dat0, dat1)
colnames(dat.wide)[1:n] <- seq(1,n)
dat.wide$id <- 1:N
```

### Version 2

Second version of data generation: ICC specified by the correlation matrix generated above -- intrasubject correlation depends on the elements of the correlation matrix

```{r Generate data V2}
set.seed(123)

datmale <- matrix(NA, nrow=N/2, ncol=n)
datfemale <- matrix(NA, nrow=N/2, ncol=n)
for (i in 1:N/2) {
  datmale[i,1:n] <- unlist(rnorm_multi(n=1, vars=n, mu=seq(1,n), sd=2, r=cmat))
  datfemale[i,1:n] <- unlist(rnorm_multi(n=1, vars=n, mu=seq(1, (n/4), length=n), sd=2, r=cmat))
}
dat.wide <- rbind(datmale, datfemale)
colnames(dat.wide)[1:n] <- seq(1,n)
gender <- c(rep(0, N/2), rep(1,N/2))
id <- seq(1:N)
dat.wide <- cbind(dat.wide, id, gender)
```

### Version 3

Third version of data generation: within subjects, each value depends on the previous value at $t-1$. This leads to more stable trends within subjects.

```{r Generate data V3, warning=FALSE}
set.seed(123)

datmale <- matrix(NA, nrow=N/2, ncol=n)
datmale[,1] <- rnorm(N/2, mean=1, sd=3)
datfemale <- matrix(NA, nrow=N/2, ncol=n)
datfemale[,1] <- rnorm(N/2, mean=1, sd=3)
for (i in 1:N/2) {
  for (j in 2:n) {
    datmale[i,j] <- rnorm(n=1, mean=datmale[i,(j-1)]+1, sd=3)
    datfemale[i,j] <- rnorm(n=1, mean=datfemale[i,(j-1)], sd=3)
  }
}
dat.wide <- rbind(datmale, datfemale)
colnames(dat.wide)[1:n] <- seq(1,n)
gender <- c(rep(0, N/2), rep(1,N/2))
id <- seq(1:N)
dat.wide <- cbind(dat.wide, id, gender)
```

After having generated the data using one of the versions above, we transform the data format from wide to long

```{r Wide to long}
sequ.tpoints <- as.character(seq(1:n))
dat <- as.data.frame(dat.wide) %>% 
  pivot_longer(cols = all_of(sequ.tpoints), names_to = "time", values_to = "y") 
dat$time <- as.numeric(dat$time)
dat$gender <- as.factor(dat$gender)
```

## Descriptives and plots

```{r Descriptives}
tapply(dat$y, dat$gender, summary) # display means per gender
boxplot(formula = y ~ gender, data = dat) # in a boxplot
```

```{r Plots}
# explore individual trajectories
ggplot(data = dat, aes(x = time, y = y, group = id, color = gender)) + geom_line()
# one line for everyone
ggplot(data = dat, aes(x = time, y = y), color = gender) + 
  geom_jitter(aes(color = gender), width = .1, height = 0) + 
  geom_smooth(method = "lm")
# base
p <- ggplot(data = dat, aes(x = time, y = y, group = id, color = gender)) + geom_jitter(width = .1, height = 0)
# different lines for gender
p  + geom_smooth(group = 0, method = "lm", data = subset(dat, dat$gender == 0)) +
     geom_smooth(group = 0, method = "lm", data = subset(dat, dat$gender == 1))

```


## Fit models

### Intercept only model

```{r Intercept only}
int.only <- lmer(y ~ 1 + (1 | id), data = dat)
summary(int.only) 
logLik(int.only)           
performance::icc(int.only) 
p + geom_smooth(method = "lm", formula = y ~ 1, se = F)
```

### Level 1 predictor: time

```{r Lvl 1}
lvl1 <- lmer(y ~ time + (1 | id), data = dat)
summary(lvl1)
logLik(lvl1)           # -1100.4 (df=4) - better fit compared to int.only
anova(int.only, lvl1)  # likelihood ratio test significant
performance::icc(lvl1) # 0.242
p + geom_smooth(method = "lm", formula = y ~ 1 + x, se = F)
```


### Level 2 predictor: gender

```{r Lvl 2}
lvl2 <- lmer(y ~ time + gender + (1 | id), data = dat)
summary(lvl2)
logLik(lvl2)           # -1063.76 (df=5) - better fit compared to lvl1 predictor only
anova(lvl1, lvl2)      # likelihood ratio test significant
performance::icc(lvl2) # 0.044
vcov.lmerMod(lvl2)
```

### Random slope for time

```{r Random slope}
rand.slop <- lmer(y ~ time + gender + (1 + time | id), data = dat)
summary(rand.slop)
logLik(rand.slop) 
anova(lvl2, rand.slop)
performance::icc(rand.slop)
```

### Interaction between time and gender

```{r Interaction effect}
inter <- lmer(y ~ time + gender + time:gender + (1 + time | id), data = dat)
summary(inter)
logLik(inter)           # -1039.767 (df=6)
anova(lvl2, inter)      # likelihood ratio test significant
performance::icc(inter) # 0.059
```



```{r}
# extract estimates, vcov matrix - Nikolas wrapper function
est <- fixef(inter)
covmat <- vcov.lmerMod(inter)
coef(inter)
result <- bain(inter, hypothesis = "time=0")
bain_2lmer(x=inter, hypotheses="time:gender=0;time:gender<0", N=100)

# Mirjams method
cat <- subset(cbind(dat$gender,TVSFP$PRETHKS),TVSFP$treat==1)
cat1[,1]=1
cat1=as.matrix(cat1)
Vmat1=Vmat[TVSFP$treat==1,TVSFP$treat==1]
Vmat1inv=solve(Vmat1)
cov1=solve(t(cat1)%*%Vmat1inv%*%cat1)


estimates <- inter$coefficients$fixed

results <- bain(estimates,hypotheses<-"Cont<CC<TV<Both;Cont<TV<CC<Both",n=ngroup,Sigma=covariances,group_parameters=1,joint_parameters = 1)

```

