---
title: "Bayesian Sample Size Estimation for Multilevel Trials"
author: "Ulrich Lösener"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

# Simulation study for the project: Bayesian Sample Size Determination for Multilevel Trials

In this simulation study, I generate multilevel data for N individuals at n timepoints. The within variable is "time" and the between variable "treatment".

First, I call the necessary libraries.

```{r Packages, message=FALSE}
library(tidyr)       # pipes
library(dplyr)       # pipes
library(ggplot2)     # plots
library(nlme)
library(lme4)        # fit multilevel model
library(lmerTest)    # get p-values of multilevel model
library(conflicted)  # conflicting commands from different packages
library(merDeriv)    # extracting vcov matrices (vcov.lmerMod)
library(mgcv)        # extracting vcov matrices
library(bain)        # Bayesian estimation
library(MASS)        # multinorm - already included in lme4?

conflicts_prefer(lme4::lmer) # tell R to prefer lme4 over lmer
conflicts_prefer(dplyr::select)
```

## Data generation

In the trajectory of treatment interventions individuals may vary both in their initial symptom level and rate of change over time. That is, people might start out the trial with different severity of their symptoms and some of them might get better while others stay the same or even get worse. Because of random allocation to treatment conditions, the initial (pre-treatment) differences in symptom levels is unrelated to the treatment groups. The further symptom trajectory of an individual, however, depends on which treatment condition they are assigned to. Therefore, I choose a model which accounts for interindividual differences at the first measurement (random intercept) that cannot be explained by the treatment condition and which allows individuals to have different rates of change over time (random slopes). These differences in slopes, in turn, can be (partly) explained by membership of a treatment condition. 

"asd"

#It is common in longitudinal intervention data that measurements closer in time correlate higher with each other than measurements further separated in time. Also, variability in the outcome typically increases over time as individuals tend to be more similar in the beginning of a study and subsequently change at different rates. This renders the assumption of compound symmetry (constant variances and covariances over time) practically untenable. To account for this, a multilevel regression model with random intercept and random slopes is used in the following simulation.  

Level 1 (within subjects) regression equation: 
$$y_{ij} = b_{0i} + b_{1i}t_j + e_{ij}$$ where $y_{ij}$ denotes the score on the outcome variable of individual i ($i=1,2,...,N$) at measurement occasion j ($j=1,2,...,n$) and $$e_{ij} \sim N(0, \sigma^2)$$

Level 2 (between subjects) regression equations: 
$$b_{0i} = \beta_0 + u_{0i}$$ 
$$b_{1i} = \beta_1 + \beta_2 X_i + u_{1i}$$ where $X_i$ denotes the treatment condition, $u_{0i}$ is the individual deviation from the treatment group intercept, and $u_{1i}$ is the individual deviation from the treatment group slope. Note that membership of the treatment group ($X_i$) does not predict initial symptom levels ($b_{0i}$) but only the rate of change over time ($b_{1i}$).

By substitution we get the total regression equation containing both levels.

$$y_{ij} = \beta_0 + u_{0i} + (\beta_1 + \beta_2 X_i + u_{1i})*t_j + e_{ij}$$
Or 

$$y_{ij} = \beta_0 + u_{0i} + \beta_1 t_j + \beta_2 X_i t_j + u_{1i} t_j + e_{ij}$$


Here, $\beta_0$ represents the average value for y at $t_0$ and $\beta_1$ the average rate of change for subjects in the $X_0$ condition. The average difference in the rate of change of y scores of subjects in the $X_0$ condition (relative to those in the $X_1$ condition) is represented by $\beta_2$. Thus, $\beta_2$ denotes the interaction between time and treatment, indicating the difference in slopes between the two treatment groups. To test whether a treatment intervention has an effect on symptom level, this is the parameter of interest. 


The population distribution of $u_{0i}$ and $u_{1i}$ is assumed to be bivariate normal $N(0, \Sigma_u)$ with 
$$\Sigma_u = \begin{bmatrix} \sigma^2_{u0} & \sigma_{u0u1}\\
\sigma_{u0u1} & \sigma^2_{u1}
\end{bmatrix}$$


```{r Data Generation}
N <- 100          # number of subjects
d <- c(0,1,2,3,4) # time of measurements
n <- length(d)    # number of measurements per subject

sigmasq.u0 <- 1  # variance of individual deviation from treatment intercept 
sigmasq.u1 <- 1  # variance of individual deviation from treatment slope
sigma.u0.u1 <- 0 # covariance between sigmasq.u0 and sigmasq.u1. If positive, then individuals with higher(lower) initial values tend to have a higher (lower) rate of change over time.
sigmasq.e <- 1 # error variance

# create data vectors
y <- rep(NA, N)  # data storage
t <- rep(d, N)
id <- rep(seq_len(N), each=n)
treat <- rep(c(0, 1), each=(N*n)/2)

beta0 <- rep(0, N*n) # average y at t0 for x=0
beta1 <- rep(1, N*n) # average increase for x=0
beta2 <- rep(.5, N*n) # average difference in slopes between conditions

#set.seed(123)      # for reproducibility

m <- 100           # number of datasets
sets <- vector("list", m)     # storage for datasets
est <- matrix(c(seq(1:m)), nrow = m, ncol = 2) # storage for fixed estimates
colnames(est) <- c("dataset", "t:treat")
bf12 <- numeric(m) # storage for BFs
bf.u <- numeric(m) # storage for BFs
bf.c <- numeric(m) # storage for BFs
pmp.b <- numeric(m) # storage for BFs

for (i in 1:m) {
multinorm <- mvrnorm(n=N,mu=c(0,0), matrix(c(sigmasq.u0, sigma.u0.u1, sigma.u0.u1, sigmasq.u1), nrow=2, ncol=2)) # draw individual deviation from treatment intercept and slope from a multivariate normal distribution with mean 0.
u0 <- rep(multinorm[,1], each=n)
u1 <- rep(multinorm[,2], each=n)
e <- rnorm(N*n, 0, sqrt(sigmasq.e))

y <- beta0 + u0 + beta1*t + beta2*treat*t + u1*t + e
dat <- data.frame(id, treat, t, y)
sets[[i]] <- dat

inter <- lme(y ~ t + t:treat, random =~ t - 1 | id, data = dat)
#, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
est[i,2] <- inter$coefficients$fixed[3]
sig <- list(as.matrix(inter$varFix[3,3]))

result <- bain(est[i,2], hypotheses <- "t:treat>0;t:treat<0", n=N, Sigma = sig, group_parameters = 1)

bf12[i] <- result$BFmatrix[1,2]
bf.u[i] <- result$fit$BF.u[1]
bf.c[i] <- result$fit$BF.c[1]
pmp.b[i] <- result$fit$PMPb[1]

}

par(mfrow=c(2,2))
plot(sort(bf.u), type="l")
#plot(sort(bf.c), type="l")
plot(sort(bf12), type="l")

hist(bf.u)
#hist(bf.c)
hist(bf12)

#hist(pmp.b)
quantile(bf12, .20)

# check the bias in estimates of beta 2
mean(est[,2]- beta2) 

```

As can be seen from the figure, individuals differ in their intercept and slope. Subjects in the treatment group ($X_1$) seem to have systematically higher slopes as compared to those in the control group ($X_0$), suggesting an interaction effect between time and treatment. Also, the variability of y scores seems to increase over time, rendering the assumption of compound symmetry untenable. 


## Descriptives/plots

```{r Descriptives, echo=F}
# plot individual trajectories
ggplot(data = dat, aes(x = t, y = y, group = id, color = as.factor(treat))) + geom_line()

boxplot(formula = y ~ treat, data = dat) # in a boxplot
```

```{r Plots, echo=F}
# one line for everyone
ggplot(data = dat, aes(x = t, y = y), color = treat) + 
  geom_jitter(aes(color = treat), width = .1, height = 0) + 
  geom_smooth(method = "lm", formula = 'y ~ x')
# base
p <- ggplot(data = dat, aes(x = t, y = y, group = id, color = treat)) + geom_jitter(width = .1, height = 0)
# different lines for treat
p  + geom_smooth(group = 0, method = "lm", data = subset(dat, dat$treat == 0), formula = 'y ~ x') + geom_smooth(group = 0, method = "lm", data = subset(dat, dat$treat == 1), formula = 'y ~ x')

```


## Fit models

### Intercept only model

```{r Intercept only}
int.only <- lmer(y ~ 1 + (1 | id), data = dat)
summary(int.only) 
logLik(int.only)           
performance::icc(int.only) 
p + geom_smooth(method = "lm", formula = y ~ 1, se = F)
```

### Level 1 (time) and level 2 (treatment) predictors plus interaction between time and treatment

```{r Interaction effect}
inter <- lme(y ~ t + treat + t:treat, random =~ 1 + time | id, data = dat)
summary(inter)
```

### Bayesian hypothesis testing about the interaction coefficient

In this section, I test whether the treatment has an effect on the symptom trajectory over time. I formulate and test three competing hypotheses about the interaction coefficient $\beta_3$ against each other, their complements and the unconstrained hypothesis $H_u$ which does not postulate any constraints on the parameter. 

$$H_1: \beta_3 > 0 \\
H_1: \beta_3 < 0 \\
H_2: \beta_3 = 0 \\
H_u: \beta_3$$


```{r Bain}

est <- inter$coefficients$fixed
#covmat <- vcov.lmerMod(inter2)
#coef(inter)

sig <- list(as.matrix(inter$varFix[4,4]))

result <- bain(est[4], hypotheses <- "time:treat>0;time:treat<0;time:treat=0", n=N*n, Sigma = sig, group_parameters = 1)
result
```

The resulting Bayes Factors tell us that $H_1$ is the hypothesis most likely to be true out of the set of considered hypotheses. We therefore conclude that in the treatment group ($X_1$), the symptom level increases more over time compared to the control group ($X_0$).


--------------------------------------------------------------------------------

## Appendix: In-between models

### Level 1 predictor: time

```{r Lvl 1}
lvl1 <- lmer(y ~ time + (1 | id), data = dat)
summary(lvl1)
logLik(lvl1)           
anova(int.only, lvl1)  
performance::icc(lvl1) 
```


### Level 2 predictor: treat

```{r Lvl 2}
lvl2 <- lmer(y ~ time + treat + (1 | id), data = dat)
summary(lvl2)
logLik(lvl2)           # -1063.76 (df=5) - better fit compared to lvl1 predictor only
anova(lvl1, lvl2)      # likelihood ratio test significant
performance::icc(lvl2) # 0.044
vcov.lmerMod(lvl2)
```

### Random slope for time

```{r Random slope}
rand.slop <- lmer(y ~ time + treat + (1 + time | id), data = dat)
summary(rand.slop)
logLik(rand.slop) 
anova(lvl2, rand.slop)
performance::icc(rand.slop)
```
