---
title: "Bayesian Sample Size Estimation for Multilevel Trials"
author: "Ulrich Lösener"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

# Bayesian Sample Size Determination for Multilevel Models

This is the R Markdown script to plot the results of the simulation study needed for the Bayesian Sample Size Determination for a multilevel model parameter. 

The parameter of interest is $\beta_2$, the coefficient of interaction between the predictors *time* and *condition.* If this coefficient is positive this indicates that people receiving the intervention get better faster over time. 

The hypotheses of interest are:
$$H_0: \beta_2 = 0$$
$$H_1: \beta_2 > 0$$

The two main functions to carry out the simulations are

1. **fct_data_generation** which generates m datasets under each hypopthesis, fits the multilevel model to each dataset and calculates m Bayes Factors for each hypothesis

2. **fct_BayesianSSD** which repeatedly applies fct_data_generation for different sample sizes (N) until the power criterion is met

First, we load the necessary packages
```{r Packages, message=FALSE, eval=FALSE}
library(tidyr)       # pipes
library(dplyr)       # pipes
library(ggplot2)     # plots
library(lme4)        # fit multilevel model
library(mgcv)        # extracting vcov matrices
library(bain)        # Bayesian estimation
library(MASS)        # multinorm - already included in lme4?
library(latex2exp)   # latex notation in plots
```

Here we plot the results of the SSD and check how different inferential tools (proportion of BFs, PMPs) behave 
```{r plots results, eval=FALSE}

# Version 1

mydat <- as.data.frame(cbind(unlist(medBF.H1), unlist(prop.BF.H1),  unlist(prop.BFc.H1), unlist(medBFc.H1), unlist(medBFu.H1), unlist(PMP.c1)))
colnames(mydat) <- c("medBF.H1", "prop.BF.H1", "prop.BFc.H1", "medBFc.H1", "medBFu.H1", "PMP.c1")
props <- as.data.frame(cbind(c(mydat[, "prop.BF.H1"], mydat[, "prop.BFc.H1"], mydat[, "PMP.c1"]), rep(c(0,1,2), each=96)))
colnames(props) <- c("BF", "type")

# prop vs. prop.c vs. PMP
props <- as.data.frame(cbind(c(mydat[, "prop.BF.H1"], mydat[, "prop.BFc.H1"], mydat[, "PMP.c1"]), rep(c(0,1,2), each=96)))
colnames(props) <- c("BF", "type")

pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/props.pmp.pdf", width = 6, height = 6)

ggplot(props) + 
  geom_line(aes(x=rep(seq(32,222, 2), 3), y=BF, colour=factor(type), linetype=factor(type)), linewidth=1) +
  xlab("N") + ylab("") + 
  scale_colour_discrete(name = "Measure of Evidence for H1", labels = c("Proportion of BF10 larger than 3", "Proportion of BFc larger than 3", "PMP")) +
  scale_linetype_discrete(name = "Measure of Evidence for H1", labels = c("Proportion of BF10 larger than 3", "Proportion of BFc larger than 3", "PMP")) +
  theme(legend.position=c(.75, 0.2))

dev.off()

# medBF vs. medBF.c
medBFs <- as.data.frame(cbind(c(mydat[, "medBF.H1"], mydat[, "medBFc.H1"]), rep(c(0,1), each=96)))
colnames(medBFs) <- c("medBF", "type")

ggplot(medBFs) + 
  geom_line(aes(x=rep(seq(32,222, 2), 2), y=medBF, colour=factor(type)), linewidth=1) +
  xlab("N") + ylab("Median BF") + xlim(c(32,180)) + ylim(c(0,100000)) +
  scale_colour_discrete(name = "Type of Measure", labels = c("Median BF10", "Median BF.c")) +
  theme(legend.position=c(.1, 0.2))


# Version 2

dat <- as.data.frame(cbind(unlist(medBF.H1), unlist(prop.BF.H1), unlist(medBFc.H1), unlist(medBFu.H1), unlist(PMP.c1)))
colnames(dat) <- c("medBF.H1", "prop.BF.H1", "medBFc.H1", "medBFu.H1", "PMP.c1")
dat <- dat[1:95,]

ggplot(dat) + 
  geom_line(aes(x=seq(32,220, 2), y=medBFc.H1, colour="red")) +
  geom_line(aes(x=seq(32,220, 2), y=medBFc.H1, colour="blue"))

# Proportion vs PMP
prop.pmp <- as.data.frame(cbind(c(dat[, "prop.BF.H1"], dat[, "PMP.c1"]), rep(c(0,1), each=95)))
colnames(prop.pmp) <- c("proppmp", "Measure")

pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/prop.pmp.pdf", width = 6, height = 6)

ggplot(prop.pmp) + 
  geom_line(aes(x=rep(seq(32,220, 2), 2), y=proppmp, colour=factor(Measure)), linewidth=1) +
  xlab("N") + ylab("") + 
  scale_colour_discrete(name = "Measure of Evidence for H1", labels = c("Proportion of BFs larger than 3", "PMP")) +
  theme(legend.position=c(.75, 0.2))

dev.off()
```

Here we examine the consequences of different effect sizes in the SSD procedure and plot the results
```{r eff size, eval=FALSE}

effsize <- rep(list(vector("list", 6)), 24)
eff <- rep(NA, 24)

for (i in 1:24) {
  effsize[i][[1]] <- dat.gen(N=100, eff.size = .2+(i/40))
  eff[i] <- .2+(i/40)
}

eff.medBF01.H0 <- rep(NA, 24)
eff.medBF10.H1 <- rep(NA, 24)
eff.propBF01.H0 <- rep(NA, 24)
eff.propBF10.H1 <- rep(NA, 24)
eff.meanBF01.H0 <- rep(NA, 24)
eff.meanBF10.H1 <- rep(NA, 24)

for(i in 1:24){
eff.medBF01.H0[i] <- effsize[[i]][1]
eff.medBF10.H1[i] <- effsize[[i]][2]
eff.propBF01.H0[i] <- effsize[[i]][3]
eff.propBF10.H1[i] <- effsize[[i]][4]
eff.meanBF01.H0[i] <- effsize[[i]][5]
eff.meanBF10.H1[i] <- effsize[[i]][6]
}

df.eff <- as.data.frame(cbind(as.numeric(eff.medBF10.H1), as.numeric(eff.propBF10.H1), as.numeric(eff.meanBF10.H1), eff))
colnames(df.eff) <- c("medBF10.H1", "propBF10.H1", "meanBF10.H1","eff.size")

mean_prop <- c(as.numeric(eff.meanBF10.H1), as.numeric(eff.propBF10.H1))
df.eff.long <- as.data.frame(cbind(mean_prop, rep(c(0,1), each=24), eff))
colnames(df.eff.long) <- c("mean_prop", "ind", "eff")

pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/effsize.pdf", width = 6, height = 6)

ggplot(data=df.eff.long, aes(x=eff, y=mean_prop, group=factor(ind), color=factor(ind))) + geom_line(linewidth=1) +
scale_color_discrete(name="", labels=c("Mean log(BF)", "Power")) +
geom_hline(yintercept = 1, linetype="dashed") +
geom_hline(yintercept = 0, linetype="dashed") +
xlab("Effect Size") + ylab("") + 
theme_grey(base_size = 22) +
theme(legend.position = c(.2, .90))

dev.off()
```

Here we summarise the results in a grid plot
```{r plots results2, eval=FALSE}
# Grid plot

ggplot(b_small, aes(x=N)) +
  geom_line(data=b_large, aes(x=N, y=prop.BF.H1), color="red", size=1) +
  geom_line(data=b_large, aes(x=N, y=PMP.c1), color="red", linetype="dashed", size=1) +

  geom_line(data=b_medium, aes(x=N, y=prop.BF.H1), color="blue", size=1) +
  geom_line(data=b_medium, aes(x=N, y=PMP.c1), color="blue", linetype="dashed", size=1) +

  geom_line(data=b_small, aes(x=N, y=prop.BF.H1), color="orange", size=1) +
  geom_line(data=b_small, aes(x=N, y=PMP.c1), color="orange", linetype="dashed", size=1) +

  #geom_line(data=b_small, aes(x=N, y=prop.BF.H0), color="black") +
  #geom_line(data=b_small, aes(x=N, y=PMP.c0), color="black", linetype="dashed") +
  
  geom_hline(yintercept=.8, col="black", linetype="dashed") 

  # labs(x = "N", y = "Proportion/PMP", color = "Legend") + scale_color_manual(values = colors)

# Plot Proportions of BF>3
pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/prop.BF10.H1.pdf", width = 6, height = 6)
ggplot(data=b, aes(x=N, y=prop.BF.H1)) + 
  geom_line(linewidth=1) +
  geom_hline(yintercept=.8, col="red", linetype="dashed") +
  labs(y="Proportion of BF01 larger than 3", x="N") #+ theme_grey(base_size = 22)
dev.off()

# Plot Median BF
pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/meanBFs.pdf", width = 6, height = 6)
ggplot(data=b) + 
  geom_line(aes(x=N, y=meanBF01.H0, color="H0"), linewidth=1) +
  geom_line(aes(x=N, y=meanBF10.H1, color="H1"), linewidth=1) +
  labs(y="Mean log(BF)", x="N", color="Correct Hypothesis") + 
  #theme_grey(base_size = 22) +
  theme(legend.position = c(.3, .90))
dev.off()
```

Here we investigate the effect of changing the number of measurement occasions per person and plot the results
```{r measurement occasions, eval=FALSE}
n.occ <- 10
meas.occ <- rep(list(vector("list", 5)), n.occ)
occ <- rep(NA, n.occ)

for(i in 1:n.occ){
  meas.occ[i][[1]] <- dat.gen.vec(m=10000, t.points = c(1:(i+2)))
  occ[i] <- (i+2)
}

occ.medBF01.H0 <- rep(NA, n.occ)
occ.medBF10.H1 <- rep(NA, n.occ)
occ.propBF01.H0 <- rep(NA, n.occ)
occ.propBF10.H1 <- rep(NA, n.occ)
occ.meanBF01.H0 <- rep(NA, n.occ)
occ.meanBF10.H1 <- rep(NA, n.occ)

for(i in 1:n.occ){
  occ.medBF01.H0[i] <- meas.occ[[i]][1]
  occ.medBF10.H1[i] <- meas.occ[[i]][2]
  occ.propBF01.H0[i] <- meas.occ[[i]][3]
  occ.propBF10.H1[i] <- meas.occ[[i]][4]
  occ.meanBF01.H0[i] <- meas.occ[[i]][5]
  occ.meanBF10.H1[i] <- meas.occ[[i]][6]
}

d2 <- as.data.frame(cbind(unlist(occ.medBF01.H0), unlist(occ.medBF10.H1), unlist(occ.propBF01.H0), unlist(occ.propBF10.H1), unlist(occ.meanBF01.H0), occ))
colnames(d2) <- c("medBF0", "medBF1", "propBF0","propBF1", "occ")

medBFs <- c(d2$medBF0, d2$medBF1)
propBFs <- c(d2$propBF0, d2$propBF1)
hyp <- rep(c(0,1), each=n.occ)
occasion <- rep(occ, 2)
d3 <- as.data.frame(cbind(medBFs, propBFs, hyp, occasion))

pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/meas.occ.propBFs.pdf", width = 6, height = 6)

ggplot(data=d3) + geom_line(aes(y=propBFs, x=occasion, group=hyp, colour=factor(hyp)), linewidth=1) + 
  xlab("Measurements per person") + ylab("Proportion of BFs larger 3") + labs(colour="Hypothesis") + scale_x_continuous(breaks=seq(1, 12))

dev.off()
```

Here we examine the effect of changing the ratio of the two variances $\sigma_u1$ and $\sigma_e$. This is inspired by chapter 9 of the book by Moerbeek and Teerenstra "Power Analysis of Trials with Multilevel Data"
```{r Function ratios, eval=FALSE}

rat <- rep(list(vector("list", 5)), 11)
slop.var <- rep(NA, 11)

for(i in 1:11){
  rat[i][[1]] <- dat.gen(var.u1 = i/10, var.e = .1, m=10000)
  slop.var[i] <- i/10
}

rats <- slop.var/.1

rat.medBF01.H0 <- rep(NA, 11)
rat.medBF10.H1 <- rep(NA, 11)
rat.propBF01.H0 <- rep(NA, 11)
rat.propBF10.H1 <- rep(NA, 11)
rat.meanBF01.H0 <- rep(NA, 11)
rat.meanBF10.H1 <- rep(NA, 11)
  
  
for(i in 1:11){
rat.medBF01.H0[i] <- rat[[i]][1]
rat.medBF10.H1[i] <- rat[[i]][2]
rat.propBF01.H0[i] <- rat[[i]][3]
rat.propBF10.H1[i] <- rat[[i]][4]
rat.meanBF01.H0[i] <- rat[[i]][5]
rat.meanBF10.H1[i] <- rat[[i]][6]
}

d <- as.data.frame(cbind(as.numeric(rat.medBF01.H0), as.numeric(rat.medBF10.H1), as.numeric(rat.propBF01.H0), as.numeric(rat.propBF10.H1),as.numeric(rat.meanBF01.H0), as.numeric(rat.meanBF10.H1), rats))
colnames(d) <- c("medBF01.H0", "medBF10.H1", "propBF01.H0","propBF10.H1","meanBF01.H0", "meanBF10.H1","ratio")

pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/medBF10.H1.pdf", width = 6, height = 6)

ggplot(data=d) + geom_line(aes(y=unlist(meanBF01.H0), x=ratio), linewidth=1) +  geom_line(aes(y=unlist(meanBF10.H1), x=ratio), linewidth=1) + xlab("ratio") + ylab("") 


dev.off()

```


Here we tried to make the SSD into a C++ Function - work in progress
```{r, eval=FALSE}
library(Rcpp)
cppFunction('NumericVector SSDc(int m, NumericVector tpoints, double var.u1, double var.e, double eff.size, double BFthres, double eta){
  int N = 30;
  double power = 0;
  Function dat.gen.cpp("dat.gen");
  
  while(power<.80){
    N = N+2;
    NumericVector results = dat.gen.cpp(m=m, N=N, t.points=t.points, ar.u0=var.u0, var.u1=var.u1, var.e=var.e, eff.size=eff.size, BFthres=BFthres);
    
    
  }
}')

```

Here we present the SSD in form of a binary algorithm in order to reduce the number of iterations and therefore save computation time. Note that this procedure cannot produce the plots above, it can only be used to determine the N necessary to meet the power criterion. 
```{r Function SSD Binary, eval=FALSE}

SSD.binary <- function(m=1000, t.points=c(1,2,3,4,5), var.u0=0, var.u1=.1, var.e=.02, eff.size=.8, BFthres=3, eta=.8){
  source("fct_data_generation.R")
  
  start <- Sys.time()
  N.current <- 30    # initial N
  N.min <- 20
  N.max <- 1000

  while (N.current!=N.min+1) {
    
    N.current <- round((N.min + N.max)/2, digits = 0)

    results <- dat.gen.vec(m=m, N=N.current, t.points=t.points, var.u0=var.u0, var.u1=var.u1, var.e=var.e, eff.size=eff.size, BFthres=BFthres)
    # condition met?
    ifelse(results$Prop_BF0>eta & results$Prop_BF1>eta, condition <- TRUE, condition <- FALSE)
    ifelse(condition==FALSE, N.min <- N.current, N.max <- N.current)
    print(N.current)
    
  }
  print(Sys.time() - start) 
  return(N.current)

}

```
  


```{r, eval=FALSE}
df.prop.bf10.H1 <- as.data.frame(cbind(prop.bf10.H1[1:length(na.exclude(ss.seq))], na.exclude(ss.seq)))
colnames(df.prop.bf10.H1) <- c("prop.bf", "N")

# increase in proportion of BFs larger than BFthres
pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/Prop.BF.medium.pdf", width = 6, height = 6)
ggplot(data=df.prop.bf10.H1, aes(x=N, y=prop.bf), linewidth=1.5) + 
  geom_line() +
  geom_hline(yintercept=eta, col="red", linetype="dashed") +
  labs(x="Proportion of BF01 larger than 3") 
dev.off()

# Sampling distribution of BFs for N
  
#bf10.H1.trim <- bf10.H1 #[bf10.H1<=quantile(bf10.H1, probs = .99)]
df.prop.bfs <- data.frame(c(bf10.H1, bf01.H0), rep(c("1","0"), each=m))
colnames(df.prop.bfs) <- c("BF", "hyp")
```

```{r, eval=FALSE}
# H0
pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/Samp.dist.H0.pdf", width = 6, height = 6)

p0 <- ggplot() + geom_density(aes(x=unlist(BFs.H0)), linewidth=1) + 
  xlim(0, 50) + 
  labs(y="density", x="BF01") 

dpb0 <- ggplot_build(p0)

p0 + geom_area(data=data.frame(x=dpb0$data[[1]]$x[dpb0$data[[1]]$x>3], y=dpb0$data[[1]]$y[dpb0$data[[1]]$x>3]), aes(x=x, y=y), fill="grey") +
    annotate("text", x=c(2,10), y=c(.01,0.005), label=c(TeX("$BF_{thresh} = 3$"), TeX("$\\eta = .80$")), angle=c(90,0)) +
    geom_vline(xintercept=3, linetype=2, color="red", linewidth=1)

dev.off()

# H1
pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/Samp.dist.H1.pdf", width = 6, height = 6)

p1 <- ggplot() + geom_density(aes(x=unlist(BFs.H1)), linewidth=1) + 
  xlim(0, 100) + 
  labs(y="density", x="BF10") 

dpb1 <- ggplot_build(p1)

p1 + geom_area(data=data.frame(x=dpb1$data[[1]]$x[dpb1$data[[1]]$x>3], y=dpb1$data[[1]]$y[dpb1$data[[1]]$x>3]), aes(x=x, y=y), fill="grey") +
    annotate("text", x=c(0,25), y=c(.01,0.0025), label=c(TeX("$BF_{thresh} = 3$"), TeX("$\\eta = .80$")), angle=c(90,0)) +
    geom_vline(xintercept=3, linetype=2, color="red", linewidth=1) 

dev.off()
```


```{r Sampling Distributions of BFs, eval=FALSE}

# both

pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/SampDistH1.pdf", width = 6, height = 6)

p + geom_area(data=data.frame(x=dpb$data[[1]]$x[dpb$data[[1]]$x>3],
                       y=dpb$data[[1]]$y[dpb$data[[1]]$x>3]),
            aes(x=x, y=y), fill="grey")
dev.off()

pdf(file = "C://Users/losen002/OneDrive - Universiteit Utrecht/Desktop/PhD/BayesianSSD/SampDistH0.pdf", width = 6, height = 6)
p0 + geom_area(data=data.frame(x=dpb0$data[[1]]$x[dpb0$data[[1]]$x>3],
                       y=dpb0$data[[1]]$y[dpb0$data[[1]]$x>3]),
            aes(x=x, y=y), fill="grey")
dev.off()

# StatAreaUnderDensity <- ggproto(
#   "StatAreaUnderDensity", Stat,
#   required_aes = "x",
#   compute_group = function(data, scales, xlim = NULL, n = 50) {
#     fun <- approxfun(density(data$x))
#     StatFunction$compute_group(data, scales, fun = fun, xlim = xlim, n = n)
#   }
# )
# 
# stat_aud <- function(mapping = NULL, data = NULL, geom = "area",
#                     position = "identity", na.rm = FALSE, show.legend = NA, 
#                     inherit.aes = TRUE, n = 50, xlim=NULL,  
#                     ...) {
#   layer(
#     stat = StatAreaUnderDensity, data = data, mapping = mapping, geom = geom, 
#     position = position, show.legend = show.legend, inherit.aes = inherit.aes,
#     params = list(xlim = xlim, n = n, ...))
# }

```

Singularity Diagnostics
```{r, eval=FALSE}
marker <- rep(NA, m)
for(i in 1:m){
  marker[i] <- ifelse(isSingular(modelsH0[[1]][[i]]), 1, 0)
}

badmodels <- modelsH0[[1]][marker==1]
goodmodels <- modelsH0[[1]][marker==0]

badest <- rep(NA, length(badmodels))
badstderr <- rep(NA, length(badmodels))
for(i in 1:length(badmodels)){
badest[i] <- badmodels[[i]]@beta[3]
badstderr[i] <- diag(vcov(badmodels[[i]]))[3]
}

goodest <- rep(NA, length(goodmodels))
goodstderr <- rep(NA, length(goodmodels))
for(i in 1:length(goodmodels)){
goodest[i] <- goodmodels[[i]]@beta[3]
goodstderr[i] <- diag(vcov(goodmodels[[i]]))[3]
}

length(goodmodels)
length(badmodels)

summary(goodest)
summary(badest)
summary(goodstderr)
summary(badstderr)
```


